Individual heterogeneity in the functional topography of the medial prefrontal cortex

### 2019-03-13 Manuscript.Rmd
Done with abstract, intro, and methods revision + updated bibliogrphy

---
title: 'Individual heterogeneity in the functional topography of the medial prefrontal cortex'
author: "Claudio Toro-Serey, Sean Tobyne, & Joseph T. McGuire"
output:
  pdf_document: 
    keep_tex: true
  word_document: 
    reference_docx: Style_reference_draft.docx
  html_document: default
bibliography: mPFC_paper.bib
header-includes:
    - \usepackage{setspace}\doublespacing
---

@@

  There is extensive topographic overlap between cortical regions associated with the default-mode network (DMN) and regions that exhibit effects of subjective value during decision making. The two sets of functional effects involve common regions of medial prefrontal cortex (mPFC) and posterior cingulate cortex (PCC). At the same time, differences between the two sets of activation patterns are not apparent at the aggregated meta-analytic level, and further individual-level functional subdivisions might be obscured by group averaging. Here we used cortical surface-based meta-analysis to identify a parcel in human PCC that was preferentially implicated in DMN effects. We then used resting-state fMRI data and a data-driven network analysis algorithm, spectral partitioning, to divide mPFC/PCC into distinct cortical communities in individual participants (n=100 from the Human Connectome Project). We labeled the communities "DMN" and "non-DMN" on the basis of their overlap with the meta-analytically identified PCC region. The spectral partitioning algorithm efficiently identified individual-level cortical subdivisions that were reliable over time and across test-retest days using moderate amounts of fMRI data (30 min per day). The topography of the identified subdivisions differed across individuals, and the degree of inter-individual variability was greater in mPFC than in PCC. Our results will aid future task-based fMRI studies in identifying functionally specific cortical subdivisions, and suggest it would be premature to infer that different task manipulations call upon shared cognitive processes solely because they evoke overlapping activation in mPFC at the group or meta-analytic level.

@@

  In this study, we aimed to subsection regions commonly attributed to both DMN and subjective value into subject-specific DMN and non-DMN partitions, quantifying the degree of topographical heterogeneity present in this network within and across individuals. We did this by capitalizing on the respective strengths of meta-analytic and subject-specific analyses of brain networks. First, we established a search space by selecting parcels from an established brain atlas [@Glasser2016] that considerably overlapped with previously-defined DMN and limbic networks along the medial wall [@Yeo2011]. A meta-analysis of DMN and SV literatures revealed areas in PCC and mPFC that were uniquely related to DMN and SV, respectively (these areas were used as markers to identify the affiliation of each individual's communities). We then produced rsFC networks of all the surface vertices within these regions of interest (ROI) for each of 100 individual resting state dataset from the Human Connectome Project (HCP)[@VanEssen2012], and estimated the extent of DMN coverage through spectral partitioning. We found that these areas of overlap can be partitioned into two communities that topographically vary across subjects, but that follow a comon organizational principle. This partition was in substantial agreement with results from a different community detection algorithm (moularity). A sliding window analysis revealed that spectral partitioning allows for a more immediate identification of noisy viertices with unreliable affiliation. Refining the partitions by filtering out these nodes resulted in idiosyncratic partitions that were highly stable across scanning runs per individual, and this pattern was higher for mPFC. We show that partitions produced by spectral partitioning had higher test-retest reliability per subject than common seed-based functional connectivity, without the need of group-drived priors [@Kong2018]. Lastly, we provide evidence that these communities resembled an organizational scheme recently observed in a handful of subjects [@Braga2017]. Our work highlights the usefulness of estimating brain effects at the individual level in these regions, and points to limitations of aggregative methods in determining the degree of neural overlap between DMN and valuation.

@@

To estimate the degree of stability of these partitions per subject we performed a sliding window analysis (20 min windows, 1 min increments, mean number of windows per subject = 37), comparing each window's resulting partition against the partition derived from the whole subject's data. A 20-min window is typically adequate for identifying stable community features in brain networks [@Gordon2017]. We assessed whether the magnitude of the Fiedler vector value for a given vertex (for the full subject-level data set) was associated with the stability of that vertex’s sub-network assignment across time windows. To do this, we fit a mixed effects logistic regression model, in which the dependent variable was the proportion of times each vertex participated in the DMN community across windows, and the explanatory variables included a random effect of subject and a fixed effect of the Fiedler vector value for that vertex (derived from their full dataset). Based on this relationship, we identified a threshold of Fiedler vector values for each subject, such that vertices above these thresholds were persistently associated with either DMN or non-DMN over 95% of the time. *Individuals without such stable nodes were not thresholded, but their maps were not discarded. Values in the eigenvector maps that did not meet threshold were set to zero, and set to 0.5 during binarization so that they would not bias the mean estimates across subjects. These thresholds yielded individual maps that retained the most stable vertices for further analyses, as well as discarding large sources of noise. (MANY NOTES: this can go into the results, and will need to add 1) % of subjects without stable nodes, and 2) the mean and range for the % of nodes that were stable.*

@@

With these `r length(unique(str_remove(Yeo_labels, "_ROI")))` ROIs (`r (length(unique(str_remove(Yeo_labels, "_ROI"))) - 2)` overlapping, 1 DMN-specific, and 1 SV-specific), we estimated each individual's DMN coverage using their full time series (4800 total TRs from four scanning sessions over two days). Each subject's network (n = `r nSubj`) was constructed by performing pairwise correlations among the time series of all the surface vertices contained within the ROIs (`r nVertices` total vertices). After correcting for multiple comparisons, we performed spectral partitioning and modularity on these networks to identify cohesive functional communities denoting DMN and non-DMN areas. On average, networks had high densities (i.e. highly connected) regardless of the amount of data to produce them (mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`). This high degree of density is important, as these methods are known to face challenges when computed on sparse networks [@Fortunato2016].

@@

``` {r Partition comparison: all 2, echo=FALSE, message = FALSE, cache = TRUE}
# Cross-method
# Don't worry about plotting, since that's mainly for a general visualization
RI_all_within <- comparePartitions(Data = Summaries, Index = "RI", MOI = c("FiedlerBinary","Membership"), nSubjects = length(Summaries), subjNames = subjList)

@@

#grid.arrange(pcc1,pfc1, ncol = 2)
#grid.arrange(ecdfALL,ecdfPCC,ecdfPFC, ncol = 1)

@@

This comparison showed that both clustering methods were in high agreement across subjects (mean = `r round(mean(RI_all_within$Index), digits = 2)` sd = `r round(sd(RI_all_within$Index), digits = 2)`). The importance of this result lies in that while spectral partitioning forces a bipartition of the network, modularity does not. This adds support to the division of these areas into DMN and non-DMN during rest. Since SP and modularity showed converging estimates, from here on we focus on results computed with spectral partitioning to ensure bipartitions.

@@

In order to address this the degree of stability of these partitions in time, we performed a sliding window analysis on each subject's full time series (20 min windows shifting at 1 min), comparing the community arrangement derived from each window with the original pattern estimated using the whole time series (Figure 4).

@@

There is precedence in thhe literature that the magnitude (and not just the sign) of the Fiedler vector values conveys important information about the role of each node in the network [@Tian2018](**add another citation, like the global map paper**). Therefore, we asked whether the magnitude of these eigenvector values was associated with the stability of nodes in time. Specifically, we examined whether Fiedler vector values could predict the proportion of DMN affiliations per node by performing a logistic mixed effects model (Figure 5). The model yielded a positive significant relationship between these features ($\beta$ = `r round(SC_FVvPDMN_modelSummary$coefficients[2,1], digits = 2)`, SE = `r round(SC_FVvPDMN_modelSummary$coefficients[2,2], digits = 2)`, *p* < 0.001), meaning that vertices with higher absolute Fiedler vector values were more persistent in their relationship with their corresponding community in time. We refer to this proportion of affiliations over time as the stability of a given node.

``` {r Threshold this stupid stuff}
# Get a FV threshold for each subject based on mean vertex stability
# Run the function safely, because some noisy subjects have no relationship between stability and FV
# Note, the mean resulting threshold is similar to the group's 25th and 75th quantiles
TFV <- safely(threshFV)
tempList <- lapply(Summaries, TFV)

@@

@@

These analyses show that the original pattern identified with the full dataset is stable in time, and that there is potential value in thresholding the Fiedler vector as a means to capture idiosyncratic and reliable DMN and non-DMN vertices, even in relatively short resting fMRI sessions (i.e. less than 20 mins). We therefore thresholded each subject's Fiedler vector to produce these refined maps. We estimated this threshold by selecting the smallest absolute Fiedler vector value that yielded a stability of 95%, for positive (mean = `r threshDescriptives[3]`, SD = `r threshDescriptives[4]`) and negative (mean = `r threshDescriptives[1]`, SD = `r threshDescriptives[2]`) values separately. Vertices that under these thresholds were set to zero (instead of removed) so that the dimensionality of the data was unaffected, and 0.5 in the binarized maps so that they wouldn't bias the calculation of averages. Figure 6 (left) shows the updated partition for the one showed in figure 2. The maps used in all subsequent analyses were thresholded by this individualized criterion.

@@

@@

With these thresholded maps, we recomputed the overall similarity of their functional organization across participants. Compared to before, there was even less overall agreement of the full maps among individuals (mean = `r round(mean(RI_all_between_lowtri), digits = 2)`, SD = `r round(sd(RI_all_between_lowtri), digits = 2)`). The same was true for both PCC (mean = `r round(mean(testPCC), digits = 2)`, SD = `r round(sd(testPCC), digits = 2)`) and mPFC (mean = `r round(mean(testPFC), digits = 2)`, SD = `r round(sd(testPFC), digits = 2)`) separately, although the significance of the differences between these areas was preserved (paired permutation, *p* = `r Overall_PCCvsPFC_permTest$Pval`; Cohen's D = `r round(Overall_PCCvsPFC_ES, digits = 2)`). Figure 6 (right) shows the spatial average of the thresholded binarized partitions across all participants, denoting the proportion of times a vertex was affiliated with the DMN community. This summary conveys a common organizational principle for both communities, but also highlights the considerable variability across individuals in mPFC compared to PCC.

@@

``` {r Partition comparison: all 2 (Fig. 4), fig.align="center", out.width = "40%", fig.cap = "Similarity matrix showing ARI values among all subjects for PCC (lower triangle) and mPFC (upper triangle) separately. PCC is better aligned across individuals than mPFC.", echo=FALSE, message = FALSE}
par(xpd=T)
corrplot(RI_PCC,
         is.corr = F, 
         method = "color",
         type = "lower", 
         tl.pos = "n", 
         tl.cex = 0.1, 
         #cl.lim = c(0,1),
         cl.length = 3,
         #addgrid.col = "grey99",
         cl.pos = "r",
         mar = c(1,1,2,0))
corrplot(RI_PFC, 
         is.corr = F, 
         method = "color", 
         type = "upper", 
         tl.pos = "n", 
         tl.cex = 0.1,
         cl.pos = "n",
         #addgrid.col = "grey90",
         mar = c(1,1,2,0), 
         add = T)
segments(0.5,0.5,0.5,100.5, lwd = 2)
segments(0.5,0.5,100.5,0.5, lwd = 2)
segments(0.5,100.5,100.5,100.5, lwd = 2)
segments(100.5,0.5,100.5,100.5, lwd = 2)
mtext(expression(bold("Subject Partitions")), line = 3)
mtext(expression(bold("Subject Partitions")), side = 2, line = -0.8)
mtext(expression(bold("mPFC")), line = 2, at = par("usr")[1]+0.73*diff(par("usr")[1:2]), col = "steelblue4")
mtext(expression(bold("PCC")), side = 1, line = 4, at = par("usr")[1]+0.18*diff(par("usr")[1:2]), col = "steelblue4")

>>>>

---
title: 'Individual heterogeneity in the functional topography of the medial prefrontal cortex'
author: "Claudio Toro-Serey, Sean Tobyne, & Joseph T. McGuire"
output:
  word_document: 
    reference_docx: Style_reference_draft.docx
  pdf_document: 
    keep_tex: true
  html_document: default
bibliography: mPFC_paper.bib
header-includes:
    - \usepackage{setspace}\doublespacing
---

@@

  There is extensive topographic overlap between cortical regions associated with the default-mode network (DMN) and regions that exhibit effects of subjective value during decision making. The two sets of functional effects involve common regions of medial prefrontal cortex (mPFC) and posterior cingulate cortex (PCC). At the same time, differences between the two sets of activation patterns are not apparent at the aggregated meta-analytic level, and further individual-level functional subdivisions might be obscured by group averaging. Here we used cortical surface-based meta-analysis to identify a parcel in human PCC that was preferentially implicated in DMN effects. We then used resting-state fMRI data and a data-driven network analysis algorithm, spectral partitioning, to divide mPFC/PCC into distinct cortical communities in individual participants (n = 100 from the Human Connectome Project). We labeled the communities "DMN" and "non-DMN" on the basis of their overlap with the meta-analytically identified PCC region. The spectral partitioning algorithm efficiently identified individual-level cortical subdivisions that were reliable over time and across test-retest days using moderate amounts of fMRI data (30 min per day). The topography of the identified subdivisions differed across individuals, and the degree of inter-individual variability was greater in mPFC than in PCC. Our results will aid future task-based fMRI studies in identifying functionally specific cortical subdivisions, and suggest it would be premature to infer that different task manipulations call upon shared cognitive processes solely because they evoke overlapping activation in mPFC at the group or meta-analytic level.

@@

  In this study, we aimed to subsection regions commonly attributed to both DMN and subjective value into subject-specific DMN and non-DMN partitions, quantifying the degree of topographical heterogeneity present in this network within and across individuals. We did this by capitalizing on the respective strengths of meta-analytic and subject-specific analyses of brain networks. First, we established a search space by selecting parcels from an established brain atlas [@Glasser2016] that considerably overlapped with previously-defined DMN and limbic networks along the medial wall [@Yeo2011]. A meta-analysis of DMN and SV literatures revealed areas in PCC and mPFC that were uniquely related to DMN and SV, respectively (these areas were used as markers to identify the affiliation of each individual's communities). We then produced rsFC networks of all the surface vertices within these regions of interest (ROI) for each of 100 individual resting state dataset from the Human Connectome Project (HCP)[@VanEssen2012], and estimated the extent of DMN coverage through spectral partitioning. We found that these areas of overlap can be partitioned into two communities that topographically vary across subjects, but that follow a comon organizational principle. This partition was in substantial agreement with results from a different community detection algorithm (modularity). A sliding window analysis showed that spectral partitioning allows for a more immediate identification of noisy viertices with unreliable affiliation. Refining the partitions by filtering out these nodes resulted in idiosyncratic partitions that were highly stable across scanning runs per individual, and this pattern was higher for mPFC. We show that partitions produced by spectral partitioning had higher test-retest reliability per subject than common seed-based functional connectivity, without the need of group-drived priors [@Kong2018]. Lastly, we provide evidence that these communities resembled an organizational scheme recently observed in a handful of subjects [@Braga2017]. Our work highlights the usefulness of estimating brain effects at the individual level in these regions, and points to limitations of aggregative methods in determining the degree of neural overlap between DMN and valuation.

@@

To estimate the degree of stability of these partitions per subject we performed a sliding window analysis (20 min windows, 1 min increments, mean number of windows per subject = 37), comparing each window's resulting partition against the partition derived from the whole subject's data. A 20-min window is typically adequate for identifying stable community features in brain networks [@Gordon2017]. We assessed whether the magnitude of the Fiedler vector value for a given vertex (for the full subject-level data set) was associated with the stability of that vertex’s sub-network assignment across time windows. To do this, we fit a mixed effects logistic regression model, in which the dependent variable was the proportion of times each vertex participated in the DMN community across windows, and the explanatory variables included a random effect of subject and a fixed effect of the Fiedler vector value for that vertex (derived from their full dataset). Based on this relationship, we identified a threshold of Fiedler vector values for each subject, such that vertices above these thresholds were persistently associated with either DMN or non-DMN over 95% of the time.

@@

With these `r length(unique(str_remove(Yeo_labels, "_ROI")))` ROIs (`r (length(unique(str_remove(Yeo_labels, "_ROI"))) - 2)` overlapping, 1 DMN-specific, and 1 SV-specific), we estimated each individual's DMN coverage using their full time series (approximately 4800 total TRs from four scanning sessions over two days). Each subject's network (n = `r nSubj`) was constructed by performing pairwise correlations among the time series of all the surface vertices contained within the ROIs (`r nVertices` total vertices). After correcting for multiple comparisons, we performed spectral partitioning and modularity on these networks to identify cohesive functional communities denoting DMN and non-DMN areas. On average, networks had high densities (i.e. highly connected) regardless of the amount of data to produce them (mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`). This high degree of density is important, as these methods are known to face challenges when computed on sparse networks [@Fortunato2016].

@@

``` {r Partition comparison: all 2, echo=FALSE, message = FALSE, cache = TRUE}
## Community number evaluation (across and within methods)
# ARI between methods
RI_all_within <- comparePartitions(Data = Summaries, Index = "RI", MOI = c("FiedlerBinary", "Membership"), nSubjects = nSubj, subjNames = subjList)

# Number of communities per subject for modularity
nModularity <- sapply(Summaries, function(data) {max(data$Membership)}) # n of estimated communities by modularity

# Median size of the extra (>2) communities estimated by modularity
modularityCounts <- lapply(Summaries, function(data) {sort(count(data$Membership)$freq, decreasing = T)})
modularityCounts_two <- sapply(modularityCounts, function(x) {mean(x[seq(2)], na.rm = T)})
modularityCounts_extra <- sapply(modularityCounts, function(x) {mean(x[3:4], na.rm = T)})

# eigenvalue-based diagnosis of optimal n of communities
prcntDissection <- lapply(Summaries, function(data) {which(data$EigenVal < 0.00001)}) # how many consecutive 0s?
prcntDissection <- mean((sapply(prcntDissection, length) + 1) == 2) * nSubj # % of subjects who only had 1 zero (thus optimal cuts = 2)

## Across subjects
# Note: the upper and lower triangles can totally be combined before plotting, instead of adding one corrplot over another. Might work on that eventually.
RI_all_between <- comparePartitions(Data = Summaries, Index = "RI", nSubjects = length(Summaries), subjNames = subjList)
RI_all_between_lowtri <- RI_all_between[lower.tri(RI_all_between)]

@@

@@

This comparison showed that both clustering methods were in high agreement across subjects (mean = `r round(mean(RI_all_within$Index), digits = 2)` sd = `r round(sd(RI_all_within$Index), digits = 2)`). Modularity showed a tendency to produce additional communities (median = `r median(nModularity)`, max = `r max(nModularity)`). However, these communities encompassed a small number of vertices (median = `r median(modularityCounts_extra, na.rm = T)`, IQR = `r IQR(modularityCounts_extra, na.rm = T)`) compared to the principal two (median = `r median(modularityCounts_two, na.rm = T)`, IQR = `r IQR(modularityCounts_two, na.rm = T)`), suggesting that the overall network is composed of two main communities. In addition, the estimation of the optimal number of cuts based on the rank of the first non-zero eigenvalue (see "Partition evaluation" for details) showed that `r prcntDissection` percent of the individual networks are best partitioned into two communities.

@@

In order to estimate the degree of stability of these partitions in time, we performed a sliding window analysis on each subject's full time series (20 min windows shifting at 1 min), comparing the community arrangement derived from each window with the partition computed using the whole time series (Figure 4).

@@

There is precedence in thhe literature that the magnitude (and not just the sign) of the Fiedler vector values conveys important information about the role of each node in the network [@Gkantsidis2003; @Tian2018]. Therefore, we asked whether the magnitude of these eigenvector values was associated with the stability of nodes in time. Specifically, we examined whether Fiedler vector values could predict the proportion of DMN affiliations per node by performing a logistic mixed effects model (Figure 5). The model yielded a positive significant relationship between these features ($\beta$ = `r round(SC_FVvPDMN_modelSummary$coefficients[2,1], digits = 2)`, SE = `r round(SC_FVvPDMN_modelSummary$coefficients[2,2], digits = 2)`, *p* < 0.001), meaning that vertices with higher absolute Fiedler vector values were more persistent in their relationship with their corresponding community in time. We refer to this proportion of affiliations over time as the stability of a given node.

``` {r Threshold values}
# Get a FV threshold for each subject based on mean vertex stability
# Run the function safely, because some noisy subjects have no relationship between stability and FV
# Note, the mean resulting threshold is similar to the group's 25th and 75th quantiles
TFV <- safely(threshFV)
tempList <- lapply(Summaries, TFV)

@@

# number of subjects without stable vertices
nUnstable <- threshDF %>% 
                filter(Lower == 0, Upper == 0) %>% 
                nrow

@@

These analyses show that the original pattern identified with the full dataset is stable in time, and that there is potential value in thresholding the Fiedler vector as a means to capture idiosyncratic and reliable DMN and non-DMN vertices, even in relatively short resting fMRI sessions (i.e. less than 20 mins). We therefore thresholded each subject's Fiedler vector to produce these refined maps. We estimated this threshold by selecting the smallest absolute Fiedler vector value that yielded a stability of 95%, for positive (mean = `r threshDescriptives[3]`, SD = `r threshDescriptives[4]`) and negative (mean = `r threshDescriptives[1]`, SD = `r threshDescriptives[2]`) values separately. Individuals without such stable nodes (n = `r nUnstable`) were not thresholded, but their partitions were not discarded. Vertices under these thresholds were set to zero (instead of removed) so that the dimensionality of the data was unaffected, and 0.5 in the binarized maps so that they wouldn't bias the calculation of averages. Figure 6 (left) shows the updated partition for the one showed in figure 2. The maps used in all subsequent analyses were thresholded by this individualized criterion.

@@

# percent of stable nodes per subject
prcntStable <- sapply(Summaries, function(data) {mean(data$FiedlerBinary != 0.5)})

@@

The average percentage of stable vertices per subject was `r round(mean(prcntStable[prcntStable < 1]), digits = 2)` (SD = `r round(sd(prcntStable[prcntStable < 1]), digits = 2)`). With these thresholded partitions, we recomputed the overall similarity of their functional organization across participants. Compared to before, there was even less overall agreement of the full maps among individuals (mean = `r round(mean(RI_all_between_lowtri), digits = 2)`, SD = `r round(sd(RI_all_between_lowtri), digits = 2)`). The same was true for both PCC (mean = `r round(mean(testPCC), digits = 2)`, SD = `r round(sd(testPCC), digits = 2)`) and mPFC (mean = `r round(mean(testPFC), digits = 2)`, SD = `r round(sd(testPFC), digits = 2)`) separately, although the significance of the differences between these areas was preserved (paired permutation, *p* = `r Overall_PCCvsPFC_permTest$Pval`; Cohen's D = `r round(Overall_PCCvsPFC_ES, digits = 2)`). Figure 6 (right) shows the spatial average of the thresholded binarized partitions across all participants, denoting the proportion of times a vertex was affiliated with the DMN community. This summary conveys a common organizational principle for both communities, but also highlights the considerable variability across individuals in mPFC compared to PCC.

@@

``` {r Partition comparison: all 2 (Fig. 4), fig.align="center", out.width = "40%", fig.cap = "Similarity matrix showing ARI values among all subjects for PCC (lower triangle) and mPFC (upper triangle) separately. PCC is better aligned across individuals than mPFC.", echo=FALSE, message = FALSE}
par(xpd = T)
corrplot(RI_PCC,
         is.corr = F, 
         method = "color",
         type = "lower", 
         tl.pos = "n", 
         tl.cex = 0.1, 
         #cl.lim = c(0,1),
         cl.length = 3,
         #addgrid.col = "grey99",
         cl.pos = "r",
         mar = c(1,1,2,0))
corrplot(RI_PFC, 
         is.corr = F, 
         method = "color", 
         type = "upper", 
         tl.pos = "n", 
         tl.cex = 0.1,
         cl.pos = "n",
         #addgrid.col = "grey90",
         mar = c(1,1,2,0), 
         add = T)
segments(0.5,0.5,0.5,100.5, lwd = 2)
segments(0.5,0.5,100.5,0.5, lwd = 2)
segments(0.5,100.5,100.5,100.5, lwd = 2)
segments(100.5,0.5,100.5,100.5, lwd = 2)
mtext(expression(bold("Subject Partitions")), line = 3)
mtext(expression(bold("Subject Partitions")), side = 2, line = -0.8)
mtext(expression(bold("mPFC")), line = 2, at = par("usr")[1]+0.73*diff(par("usr")[1:2]), col = "steelblue4")
mtext(expression(bold("PCC")), side = 1, line = 4, at = par("usr")[1]+0.18*diff(par("usr")[1:2]), col = "steelblue4")
